data-exploration

### 2022-04-24 outputs/paper.Rmd
Creating plots in paper file

---
title: "TTC delay times are not random"
subtitle: "How long you wait depends on where you are, what route you take, and when you take it"
author: 
  - Hudson Yuen
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Despite ranking amongst North America’s best transit systems, the Toronto Transit Commission (TTC) is often derided for irritating delays and service interruptions. This report examines TTC bus, subway, and streetcar delay data, aiming to identify the factors influencing such delays and how improvements can be made. Certain incident types and days of service see disproportionately long delays, creating opportunities to address the most unpleasant aspects of the rider experience in a targeted manner. Doing so will not only strengthen the city’s ability to move people, but to also facilitate sustainable development. **Keywords:** public transit, transit efficiency, urban mobility, sustainble transportation, transportation equity"
output:
  bookdown::pdf_document2
toc: TRUE
thanks: "All data and scripts available at https://github.com/hudyu17/transit-delays"
fig_caption: yes
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(data.table)
library(dplyr)
# library(kableExtra)
```
\newpage

@@

The future competitiveness and environmental sustainability of Toronto is intrinsically tied to the strength of the TTC. An inadequate transit system will hamstring growth and opportunity, both on individual scales and beyond [@citeTransit]. Understanding where delays are the most severe can not only help address current limitations, but also aid prediction and prevention of future bottlenecks as projects like the Ontario Line begin construction through the downtown core.

@@

\newpage
# Data

@@

## Data Characteristics

@@

>>>>

---
title: "TTC bus delay times are not random"
subtitle: "How long you wait depends on where you are, what route you take, and when you take it"
author: 
  - Hudson Yuen
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Despite ranking amongst North America’s best transit systems, the TTC is often derided for irritating delays and service interruptions. This report examines TTC bus delay data, aiming to identify the factors influencing such delays and how improvements can be made. Certain incident types and days of service see disproportionately long delays, creating opportunities to address the most unpleasant aspects of the rider experience in a targeted manner. Doing so will not only strengthen the city’s ability to move people, but to also facilitate sustainable development. **Keywords:** public transit, transit efficiency, urban mobility, sustainble transportation, transportation equity"
output:
  bookdown::pdf_document2
toc: TRUE
thanks: "All data and scripts available at https://github.com/hudyu17/transit-delays"
fig_caption: yes
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(data.table)
library(dplyr)
library(gridExtra)
# library(kableExtra)

library(rcompanion)
library(AICcmodavg)
library(Rmisc)
```
\newpage

@@

The future competitiveness and environmental sustainability of Toronto is intrinsically tied to the strength of the TTC. An inadequate transit system will hamstring growth and opportunity, both on individual scales and beyond [@citeTransit]. Buses will remain a critical aspect of our transit network due to their relatively low cost of implementation and high operational flexibility, especially in the sprawling Greater Toronto Area. Understanding where delays are the most severe can not only help address current limitations, but also aid prediction and prevention of future bottlenecks as projects like the Ontario Line begin construction through the downtown core.

@@

\newpage
# Data
```{r}
df <- readRDS('../inputs/bus-clean.RDS')

df[, month_year := format(as.Date(df$report_date), "%Y-%m")]

df[, year := format(as.Date(df$report_date), "%Y")]

incidents <- unique(df[, incident])
df[incident %like% "^Late Leaving Garage", incident := "Late Leaving Garage"]
df[incident %like% "^Road Block", incident := "Road Block"]
df[incident %like% "^Cleaning", incident := "Cleaning"]
df[incident %like% "^Late Entering Service", incident := "Late Entering Service"]
df[incident %like% "^Operations", incident := "Operations"]
df[incident %like% "^Collision", incident := "Collision"]
df[incident %like% "^Late Leaving Garage", incident := "Late Leaving Garage"]
df[incident == "Securitty", incident := "Security"]
df[incident == "Utilized Off Route", incident := "Diversion"]
df[incident %in% c("Held By", "Late", "Overhead"), incident := "General Delay"]
```

@@

## Data Characteristics
```{r, fig.cap="Delay behaviour over time", out.width="70%", fig.align='center'}
df[, mean(min_delay), by=month_year] %>%
    ggplot(aes(x=month_year, y=V1, group=1)) +
    geom_line() + 
    ylab("Average monthly delay (minutes)") + 
    xlab("Time (2014-2021)") +
    ggtitle("Bus delay duration over time") + 
    theme_minimal() + 
    theme(  
      axis.text.x = element_blank(),
      axis.ticks = element_blank(),
      panel.grid.major = element_blank()
    )
```

```{r, fig.cap="Delay behaviour by day of week", out.width="90%", fig.align='center'}
df_day <- copy(df[, mean(min_delay), by=day])
df_day$day <- factor(df_day$day, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))

p_day_duration <- df_day %>%
    ggplot(aes(x=day, y=V1)) + 
    geom_bar(stat="identity", width=0.7) +
    ylab("Average daily delay (minutes)") + 
    xlab("Day") +
    ggtitle("Delay duration") + 
    theme_minimal() + 
    theme(  
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle=90, hjust=1)
    )

df_day_count <- copy(df[, .N, by=day])
df_day_count$day <- factor(df_day$day, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))

p_day_count <- df_day_count %>%
  ggplot(aes(x = day, y = N)) + 
	geom_bar(stat = 'identity', width=0.7) + 
  ylab("Count") + 
  ggtitle("Number of delays") + 
	theme_minimal() + 
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle=90, hjust=1)
  )

grid.arrange(p_day_duration, p_day_count, nrow = 1)
```
```{r, fig.cap="Delay behaviour by incident type", out.width="90%", fig.align='center'}
p_incident_duration <- df[, mean(min_delay), by=incident] %>%
    ggplot(aes(x=reorder(incident, -V1), y=V1)) + 
    geom_bar(stat="identity", width=0.7) + 
    xlab("Incident") +
    ylab("Average delay (minutes)") + 
    ggtitle("Delay duration") + 
    theme_minimal() +
    theme(axis.text.x = element_text(angle=90, hjust=1))

p_incident_count <- df[,.(count = .N), by = incident] %>%
  ggplot(aes(x = reorder(incident,(-count)), y = count)) + 
	geom_bar(stat = 'identity', width=0.7) + 
  xlab("Incident") +
  ylab("Count") + 
  ggtitle("Number of delays") + 
	theme_minimal() + 
  theme(axis.text.x = element_text(angle=90, hjust=1))

grid.arrange(p_incident_duration, p_incident_count, nrow = 1)
```

@@

## Goals

## ANOVA

### Data Transformation
```{r, fig.cap="Histogram of delay durations", out.width="70%", fig.align='center'}
df %>%
  ggplot(aes(x=min_delay)) + 
  geom_histogram(bins=50) + 
  ylab("Count") + 
  xlab("Delay duration") +
  ggtitle("Count of delays by duration") + 
	theme_minimal() +
  theme(
    axis.ticks.y = element_blank()
  )
```
```{r histogram, fig.cap="Histogram of delays after log transformation", out.width="70%", fig.align='center'}
delay <- df[, min_delay]
delay_log = log(delay)
delay_log <- delay_log[!is.infinite(delay_log)]
plotNormalHistogram(delay_log)

df[, delay_log := log(min_delay)]
df <- df[!is.infinite(delay_log)]
```

### Model and Feature Selection
```{r}
df[, hour := substr(time, 1, 2)] # keep as character because categorical

```

### Post-hoc Tests

### Follow up: Trends Over Time

## Limitations

### 2022-04-24 scripts/data-exploration.Rmd
Creating plots in paper file

library(tidyverse)
library(data.table)
library(lubridate)
library(rcompanion)
library(AICcmodavg)
```

@@

# Summary stats
## Delay over time
```{r}
df[, month_year := format(as.Date(df$report_date), "%Y-%m")]

@@

incidents <- unique(df[, incident])
for (curr_incident in incidents) {
  p <- df[incident == curr_incident, mean(min_delay), by=month_year] %>%
    ggplot(aes(x=month_year, y=V1, group=1)) +
    geom_line() +
    ggtitle(curr_incident)

@@

```
## Delay by day of week
```{r}
df[incident == 'Mechanical', mean(min_delay), by=day] %>%
    ggplot(aes(x= reorder(day,-V1), y=V1)) + 
    geom_bar(stat="identity") 
```
# Quick modelling
## Check distribution of delays
```{r}
# exploration, no need to run
df %>%
  ggplot(aes(x=min_delay)) + 
  geom_histogram()

@@

two_way_day_incident <- aov(delay_log ~ day + incident, data = df2015)
two_way_day_hour <- aov(delay_log ~ day + hour, data = df2015)
two_way_incident_hour <- aov(delay_log ~ hour + incident, data = df2015)

@@

# anova for delay over time using month-year
```{r}
month_year_aov <- aov(delay_log ~ month_year, data = df)
summary(month_year_aov)

@@

year_model <- lm(min_delay ~ year, data = df)
summary(year_model)

df[, mean(min_delay), by=year] %>%
    ggplot(aes(x=year, y=V1, group=1)) +
    geom_line() 
```

>>>>

library(tidyverse)
library(data.table)
library(lubridate)
library(rcompanion)
library(AICcmodavg)
library(Rmisc)
```

@@

# some more cleaning of the incident type
```{r}
incidents <- unique(df[, incident])
incidents

df[incident %like% "^Late Leaving Garage", incident := "Late Leaving Garage"]
df[incident %like% "^Road Block", incident := "Road Block"]
df[incident %like% "^Cleaning", incident := "Cleaning"]
df[incident %like% "^Late Entering Service", incident := "Late Entering Service"]
df[incident %like% "^Operations", incident := "Operations"]
df[incident %like% "^Collision", incident := "Collision"]
df[incident %like% "^Late Leaving Garage", incident := "Late Leaving Garage"]
df[incident == "Securitty", incident := "Security"]
df[incident == "Utilized Off Route", incident := "Diversion"]
df[incident %in% c("Held By", "Late", "Overhead"), incident := "General Delay"]
```

# Summary stats / data characteristics
## Delay over time
```{r}
df[, month_year := format(as.Date(df$report_date), "%Y-%m")]

@@

for (curr_incident in incidents) {
  p <- df[incident == curr_incident, mean(min_delay), by=month_year] %>%
    ggplot(aes(x=month_year, y=V1, group=1)) +
    geom_line() +
    ggtitle(curr_incident)

@@

```

## Delay by day of week
```{r}
# TODO: reorder by day of week
df[, mean(min_delay), by=day] %>%
    ggplot(aes(x= reorder(day,c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday')), y=V1)) + 
    geom_bar(stat="identity") 
```
# Quick modelling
## Check distribution of delays
```{r}
# exploration, no need to run
df %>%
  ggplot(aes(x=min_delay)) + 
  geom_histogram()

@@

two_way_day_incident <- aov(delay_log ~ day + incident, data = df)
two_way_day_hour <- aov(delay_log ~ day + hour, data = df2015)
two_way_incident_hour <- aov(delay_log ~ hour + incident, data = df2015)

@@

# plots for results
```{r}
sumstats = summarySE(df,
                measurevar="min_delay",
                groupvars=c("incident"))
sumstats

pd = position_dodge(.2)

sumstats %>% ggplot(aes(x=factor(incident), y=min_delay)) + 
  geom_bar(stat="identity", fill="blue", width=0.8) +
  geom_errorbar(aes(ymin=min_delay-se, ymax=min_delay+se), width=.2, colour='red') +
  xlab("Incident") +
  ylab("Delay (minutes)") + 
  theme(axis.text.x = element_text(angle=90, hjust=1))

df %>% ggplot(aes(x = incident, y = min_delay)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

## hour plots
```{r}
df %>%
  ggplot(aes(x=hour, y=min_delay)) + 
  # geom_errorbar(aes(ymin=min_delay-se, ymax=min_delay+se), width=.2, colour='red') +
  stat_summary(fun.y = "mean", geom = "bar")

df %>% ggplot(aes(x = hour, y = min_delay)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

## day plot
```{r}
df %>% ggplot(aes(x = day, y = min_delay)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

## anova for delay over time using month-year
```{r}
month_year_aov <- aov(delay_log ~ month_year, data = df)
summary(month_year_aov)

@@

year_model <- lm(V1 ~ as.numeric(year), data = df[, mean(min_delay), by=year])
summary(year_model)

df[, mean(min_delay), by=year] %>%
    ggplot(aes(x=year, y=V1, group=1)) +
    geom_line() + 
    geom_smooth(method='lm', formula= y~x)

df[incident=='Mechanical', mean(min_delay), by=year] %>%
    ggplot(aes(x=year, y=V1, group=1)) +
    geom_line() + 
    geom_smooth(method='lm', formula= y~x)

df[incident=='Cleaning', mean(min_delay), by=year] %>%
    ggplot(aes(x=year, y=V1, group=1)) +
    geom_line() +
    geom_smooth(method='lm', formula= y~x)

# show some were flat e.g. late leaving garage flat until 2021
year_aov_llg <- aov(delay_log ~ year, data = df[incident=='Late Leaving Garage' & year != 2021])
summary(year_aov_llg)

df[incident=='Late Leaving Garage' & year != 2021] %>% ggplot(aes(x = year, y = min_delay)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle=90, hjust=1))

df[incident=='Late Leaving Garage', mean(min_delay), by=year] %>%
    ggplot(aes(x=year, y=V1, group=1)) +
    geom_line()

TukeyHSD(year_aov_llg)
```

## worst routes
```{r}
df[, mean(min_delay), by=route][order(-V1)][1:100]

# worst delayed
df[, mean(min_delay), by=route][order(-V1)][1:25] %>%
    ggplot(aes(x= reorder(route,-V1), y=V1)) + 
    geom_bar(stat="identity") # are these night buses?

# least delayed
df[, mean(min_delay), by=route][order(V1)][1:30] %>%
    ggplot(aes(x= reorder(route,V1), y=V1)) + 
    geom_bar(stat="identity") # are these night buses?

df[, mean(min_delay), by=route] %>%
    ggplot(aes(x= reorder(route,-V1), y=V1)) + 
    geom_bar(stat="identity") 
```
